Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-04 manuscript/manuscript.tex
updated ms

\vb{One concern is that these covariances reflect a few large-effect loci
  rather than selection on a polygenic trait. Since our covariances are
  essentially averages over loci, the covariance estimate can be strongly
  affected by a single outlier region. To test whether large outlier regions
  drive the genome-wide signal we see in the \textcite{Barghi2019-qy} data, we
  calculate the covariances in 100kb windows along the genome (we refer to
  these as windowed covariances throughout) and exclude the upper and lower
  $\alpha$ tails to calculate the trimmed average windowed covariance. As
  $\alpha$ increases to 0.5, the trimmed covariance converges to the median
  windowed covariance (by the definition of the median; see Supplementary
  Figure \ref{suppfig:barghi-trimmed-mean}). In other words, this median
  windowed covariance is a measure of the genome-wide covariance robust to
  large-effect loci; we find that the median windowed covariance $\cov(\Delta
  p_0, \Delta p_{10})$ (pooling all replicates) is also significantly different
  from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017]).

@@

Second, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Castro2019-uk,Marchini2014-de}. This study includes two independent selection lines,
Longshanks 1 and 2 (LS1 and LS2), and an unselected control line (Ctrl).
Consequently, this selection experiment offers a useful control to test our
between-replicate covariances: we expect to see positive between-replicate
covariance in the comparison between the two Longshanks selection lines, but
not between the two pairwise comparisons between the control line and the two
Longshanks lines. We find that this is the case (gray confidence intervals in
Figure \ref{fig:figure-2} C), with the two Longshanks comparisons to the
control line not being significantly different from zero, while the comparison
between the two Longshanks line is statistically significantly different from
zero (CIs [0.013, 0.037]).

@@

\begin{align}
  \label{eq:}
  \Delta p_t &= \underbrace{\Delta_{_N} p_t  + \Delta_{_M} p_t}_\text{drift} + \underbrace{\Delta_{_S} p_t}_\text{shared} + \underbrace{\Delta_{_U} p_t}_\text{unique}
\end{align}

@@

where $\E_{A \ne B}$ indicates that the expectation is taken over all ordered
pairs of replicates (e.g. summing all off-diagonal elements replicate
covariances), and $\E_R$ indicates taking expectation over all replicates.

>>>>

\vb{One concern is that these covariances reflect a few large-effect loci
  rather than selection on a polygenic trait. Since our covariances are
  essentially averages over loci, the covariance estimate can be strongly
  affected by a single outlier region. To test whether large outlier regions
  drive the genome-wide signal we see in the \textcite{Barghi2019-qy} data, we
  calculate the covariances in 100kb windows along the genome (we refer to
  these as windowed covariances throughout) and exclude the upper and lower
  $\alpha$ tails to calculate the trimmed average windowed covariance. As
  $\alpha$ increases to 0.5, the trimmed covariance converges to the median
  windowed covariance (by the definition of the median; see Supplementary
  Figure \ref{suppfig:barghi-trimmed-mean}). In other words, this median
  windowed covariance is a measure of the genome-wide covariance robust to
  large-effect loci; we find that the median windowed covariance $\cov(\Delta
  p_0, \Delta p_{10})$ (pooling all replicates) is also significantly different
  from zero, 0.0016 (0.95\% CIS [0.0015, 0.0017]). 
 }

@@

Second, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Castro2019-uk,Marchini2014-de}. This study includes two independent
selection lines, Longshanks 1 and 2 (LS1 and LS2), and an unselected control
line (Ctrl).  Consequently, this selection experiment offers a useful control
to test our between-replicate covariances: we expect to see positive
between-replicate covariance in the comparison between the two Longshanks
selection lines, but not between the two pairwise comparisons between the
control line and the two Longshanks lines. We find that this is the case (gray
confidence intervals in Figure \ref{fig:figure-2} C), with the two Longshanks
comparisons to the control line not being significantly different from zero,
while the comparison between the two Longshanks lines is statistically
significantly different from zero (CIs [0.013, 0.037]).

\vb{The presence of an unselected control line provides an alternative way to
  partition the effects of linked selection and genetic drift.

@@

\begin{align}
  \label{eq:shared-partition}
  \Delta p_t &= \underbrace{\Delta_{_N} p_t  + \Delta_{_M} p_t}_\text{drift} + \underbrace{\Delta_{_S} p_t}_\text{shared} + \underbrace{\Delta_{_U} p_t}_\text{unique}
\end{align}

@@

where $\E_{A \ne B}$ indicates that the expectation is taken over all ordered
pairs of replicates (e.g. summing all off-diagonal elements replicate
covariances), and $\E_R$ indicates taking expectation over all replicates. This
measures the fraction of variance in allele frequency change (averaged across
replicates) due to shared selection pressure.

For selection studies with a control line (e.g. the \cite{Castro2019-uk} study
we have reanalyzed), we have an alternate approach to quantify the effect of
linked selection: we can compare the total variance in allele frequency change
in control and selection lines. By the orthogonality of the components in
equation XXX,

\begin{align}
  \var(\Delta p_t - p_0) = \sum_t \var(\Delta {_N} p_t ) + \sum_t \var(\Delta {_M} p_t ) + 
  \var\left(\sum_t \Delta {_S} p_t\right) + \var\left(\sum_t \Delta {_U} p_t\right)
\end{align}

the term $\sum_t \var(\Delta {_S} p_t)$

\begin{align}
  \var\left(\sum_t \Delta_{_S} p_t\right) = 
 \sum_t \var( \Delta_{_U} p_t) + \sum_{i \ne j} \cov( \Delta_{_S} p_i, \Delta_{_S} p_j)
\end{align}

In the \textcite{Castro2019-uk} study, we 
quantify the effect of linked selection as:

\begin{align}
  \frac{1}{2} \left( \var(\Delta p_{20, \mathrm{LS1}} - p_{0, \mathrm{LS1}}) + \var(\Delta p_{20, \mathrm{LS2}} - p_{0, \mathrm{LS2}}) \right) -  \var(\Delta p_{20, \mathrm{Ctrl}} - p_{0, \mathrm{Ctrl}}) = \\
  \cov(p_{20, \mathrm{LS1}} - p_{0, \mathrm{LS1}}, p_{20, \mathrm{LS2}} - p_{0, \mathrm{LS2}})
\end{align}

% \begin{align}
%   \var(p_t - p_0) &= \sum_i \var(\Delta p_i) + \sum_{i\ne j} \cov(\Delta p_i, \Delta p_j) \\
%                   &= \sum_i \var(\Delta p_i) + \sum_{i\ne j} \cov(\Delta_{_U} p_i, \Delta_{_U} p_j) + \sum_{i\ne j} \cov(\Delta_{_S} p_i, \Delta_{_S} p_j)  \\
% \end{align}

% However, with only two timepoints, we cannot directly measure the unique
% temporal covariance. However, we can directly measure the covariance between
% replicates $A$ and $B$, $\sum_{i,j} \cov(\Delta p_{i,A}, \Delta p_{j,B})$.
